---
title: "03_project_environments"
output: html_document
---

layout: true

---

class: middle, center, inverse

# Project Environments

---

## 5 `renv`: Project Environments
### 5.1 Library Paths

**Question:** Where does R store my installed packages respectively loads them from?
--
<br><br>
**Answer:** Your user-specific *active library path*, i.e. on your disk.
--
<br>
```
.libPaths()
```
```
> [1] "C:/Users/ziv_kennung/Documents/R/R-4.0.2/library"
```
--
<br>
Packages are stored and loaded from the root installation folder (here for Windows operating systems). R crawls this path when loading (`library()`) or downloading packages (`install.packages()`).
```
find.package("splines")
```
```
> [1] "C:/Users/.../Documents/R/R-4.0.2/library/splines"
```

???
comment

---

## 5 `renv`: Project Environments
### 5.1 Library Paths

By default, this path is accessed each time you run a new `R` session or create a new `.Rproj`. Yet, the required version of a given package may vary across projects (e.g., depending on the project's start date):

- Project A requires `dplyr 1.0.0`,
- Project B requires `dplyr 0.8.1`, and
- Project C requires the development version of `dplyr`.

According to what we have learned, the three projects share the same mutual library path (`C:/Users/ziv_kennung/Documents/R/R-4.0.2/library`). An upgrade or downgrade of the package version increases the chance that one of the projects does not work seamlessly any longer.

???
comment

---

background-image: url(https://rstudio.github.io/renv/reference/figures/logo.svg)
background-position: 95% 5%
background-size: 7.5%
layout: true

---

## 5 `renv`: Project Environments
### 5.2 Features of `renv`

`R` package for initializing and managing local project environments, i.e. collections of R packages utilized within an *.Rproj*.

**Isolated projects:**<br>
- One local library of R packages per `.Rproj`
- Upgrade/downgrade packages without breaking other projects

**Portable projects:**<br>
- Save the state of your project library in a `lockfile`
- Share `lockfile` with others and ensure library compatibility

**Reproducible projects:**<br>
- Use `renv::snapshot()` to save the state of your project library
- Use `use renv::restore()` to restore the state of your project library as specified in the lockfile

???
comment

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow

**Step 1:**  
Set up `renv` by running `renv::init()` to initialize your local environment with a private library of `R` packages specific to your project.

**Step 2:**  
Write some code, install, uninstall and update some packages...

**Step 3:**  
Use `renv::snapshot()` to save the current state of your project library to a so-called *lockfile*.

**Step 4:**  
Write some code, install, uninstall and update some packages...

**Step 5:**  
Use `renv::restore()` to revert to the library state encoded in your *lockfile* in case you run into problems while working with updated packages.

???
comment

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv::init()`

```
> install.packages("renv")
> renv::init()
```
--
<br>
Under the hood, `renv::init()` establishes the infrastructure to run your project-specific library. More specifically, the following files are written to and used by your `renv`-enabled projects:
<br>
```{r, echo=F}
tbl <- data.frame(
  File = c(".Rprofile", "renv.lock", "renv/activate.R", "renv/library"),
  Usage = c("Used to activate renv whenever a new R session is launched within the project.", "The lockfile, defining the state of your project library.", "The activation script run by the project .Rprofile.", "Your local project library.")
)

knitr::kable(tbl, format = 'html')
```

???
comment

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv.lock`

The **lockfile** stores all information necessary to re-install the required packages, e.g., when sharing the project with a collaborator, archiving a project or putting code into productive use (*deployment*). Packages can be downloaded from: `CRAN`, `Bioconductor`, `GitHub`, `Gitlab`, `Bitbucket`.

**Exemplary lockfile entry in JSON format:**
```
"renv": {
  "Package": "renv",
  "Version": "0.11.0",
  "Source": "Repository",
  "Repository": "CRAN",
  "Hash": "1c3ef87cbb81c23ac96797781ec7aecc"
}
```
???
JSON: list of key-value pairs
you cultivate on lock-file per project. as soon as you feel you need to require another lockfile (e.g., due to differeing package versions) you should rather start a new R project.

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv/library`

`renv` links your project to a **global package cache** where all your downloaded packages are stored by `renv`. You can think of it as a repository that contains all package versions ever installed on your machine.

.pull-left[
```{r, echo=F, out.width='100%', out.height='100%'}
knitr::include_graphics("https://kevinushey-2020-rstudio-conf.netlify.app/img/package-cache.svg")
```
]
.pull-right[
This approach allows you to avoid redundancies, if the same package version is used in multiple projects.

In this case, the installation of the respective package is stored in the global cache with `renv` simply creating a link to this cache within your project. This is what your local **project library** under the path `renv/library` is all about.
]
.pull-right[
.footnote[
Picture by [Kevin Ushey](https://kevinushey-2020-rstudio-conf.netlify.app/slides.html#1).
]]

???
- this approach saves space on disk and makes installing and restoring packages much faster
- renv only copies packages if the linking approach fails

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv/library`

Finally, if we now check our active library path (`.libPaths()`), we observe a new project-specific path (as well as an additional system library path which you need to consider further at this point).
```{r}
.libPaths()
```

???
comments

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv::snapshot()`

After having successfully set up our project-specific library, let's consider the two other main features horses of the `renv` package: `renv::snapshot()` and `renv::restore()`

**Snapshot:** updates the lockfile by writing the current state of the project library to the lockfile.

```{r, eval=F}
renv::snapshot()
```
--
```
> The following package(s) will be updated in the lockfile:
> 
> # CRAN ===============================
> - glue        [1.4.0 -> 1.4.1]
> - jsonlite    [1.6.0 -> 1.7.0]
> - rmarkdown   [* -> 2.3]
> 
> Do you want to proceed? [y/N]: y
>  * Lockfile written to 'C:/Users/s_scho53/Desktop/tidyverse workshop/renv.lock'.
```

???
comment

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv::restore()`

After having successfully set up our project-specific library, let's consider the two other main features horses of the `renv` package: `renv::snapshot()` and `renv::restore()`

**Restore:** updates the project library by restoring the state of the project library as specified in the lockfile.

```{r, eval=F}
renv::restore()
```
--
```
> The following package(s) will be updated:
> 
> # CRAN ===============================
> - rmarkdown   [* -> 2.3]
> - stringr     [* -> 1.4.0]
> 
> Do you want to proceed? [y/N]: y
> Installing stringr [1.4.0] ... OK (copied cache)
> Installing rmarkdown [2.3] ... OK (copied cache)
```

???
comment

---

## 5 `renv`: Project Environments
### 5.4 Additional Remarks

- A project is always bound to a particular version of `renv`. Updating the current version of the `renv` package should be done using the `renv::upgrade()` command.<br><br>
- When setting up `renv` for an existing project, upon calling `renv::init()` `renv` autonomously crawls all project scripts and searches for packages that are mentioned in your code via `renv::dependencies()`. For this presentation, I relied - among others - on the following packages (note that packages are only identified if explicitly mentioned in the code, i.e. by name):
```{r, eval=F}
head(renv::dependencies(), 5)
```
```{r, echo=F}
df <- renv::dependencies()

df$Source <- sub(
  pattern = "Desktop/L09_Desktop/workshops/01_structuring_data_projects",
  replacement = "...",
  df$Source)

head(df, 5)[,c("Source", "Package", "Version")]
```

???
Here, only the top 5 entries. `knitr` is three times, according by how often I use the package in my code (here three times for including pictures and gifs in the slides).

