---
title: "03_project_environments"
output: html_document
---

layout: true

---

class: middle, center, inverse

# 5 Project Environments

---

## 5 `renv`: Project Environments
### 5.1 Library Paths

**Question:** Where does `R` generally store my installed packages respectively loads them from?
--
<br><br>
**Answer:** Your user-specific *active library path*, i.e. from disk.
--

```{r, eval=F}
.libPaths()
```
```
> [1] "C:/Users/.../Documents/R/R-4.0.2/library"
```
--

Packages are stored and loaded from the root installation folder (here for Windows operating systems). `R` crawls this path when loading (`library()`) or downloading (`install.packages()`) packages.
```{r, eval=F}
find.package("splines")
```
```
> [1] "C:/Users/.../Documents/R/R-4.0.2/library/splines"
```

---

## 5 `renv`: Project Environments
### 5.1 Library Paths

By default, this path is accessed each time you run a new `R` session or create a new `.Rproj`. Yet, the required version of a given package may vary across projects (e.g., depending on the project's start date):

- Project A requires `dplyr 1.0.0`,
- Project B requires `dplyr 0.8.1`, and
- Project C requires the dev version of `dplyr`.

According to what you have learned, the three projects share the same mutual library path:
```
C:/Users/.../Documents/R/R-4.0.2/library
```
An upgrade or downgrade of the package version increases the chance that one of the projects does not work seamlessly any longer.

---

background-image: url(https://rstudio.github.io/renv/reference/figures/logo.svg)
background-position: 95% 5%
background-size: 7.5%
layout: true

---

## 5 `renv`: Project Environments
### 5.2 Features of `renv`

`R` package for initializing and managing local project environments, i.e. collections of R packages utilized only within a certain *.Rproj*.

**Isolated projects:**  
- One local library of R packages per `.Rproj`
- Upgrade/downgrade packages without breaking other projects

**Portable projects:**  
- Save the state of your project library in a `lockfile`
- Share the `lockfile` with others and ensure compatibility

**Reproducible projects:**  
- Use `renv::snapshot()` to save the state of your project library
- Use `renv::restore()` to restore the state of your project library as specified in the lockfile

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow

**Step 1:**  
Set up `renv` by running `renv::init()` to initialize your local environment with a private library of `R` packages specific to your project.

**Step 2:**  
Write some code, install, uninstall and update some packages.

**Step 3:**  
Use `renv::snapshot()` to writte the current state of your project library to the so-called *lockfile*.

**Step 4:**  
Write some code, install, uninstall and update some packages.

**Step 5:**  
Use `renv::restore()` to revert to the library state encoded in your *lockfile* in case you run into problems while working with updated packages.

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv::init()`

```
> install.packages("renv")
> renv::init()
```

Under the hood, `renv::init()` establishes the infrastructure to run your project-specific library. More specifically, the following files are written to and used by your `renv`-enabled projects:

```{r, echo=F}
tbl <- data.frame(
  File = c(".Rprofile", "renv.lock", "renv/activate.R", "renv/library"),
  Usage = c("Used to activate renv whenever a new R session is launched within the project", "The lockfile, defining the state of your project library", "The activation script run by the project .Rprofile", "Your local project library")
)

knitr::kable(tbl, format = 'html')
```

???
this is not super important to know but you might wander what all the new files in your project folder are meant for

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv.lock`

The **lockfile** stores all information necessary to re-install the required packages, e.g., when sharing the project with a collaborator, archiving a project or putting code into productive use (*deployment*).

Packages can be downloaded from: `CRAN`, `Bioconductor`, `GitHub`, `Gitlab`, `Bitbucket`.

--

**Exemplary lockfile entry in JSON format:**
```
"renv": {
  "Package": "renv",
  "Version": "0.11.0",
  "Source": "Repository",
  "Repository": "CRAN",
  "Hash": "1c3ef87cbb81c23ac96797781ec7aecc"
}
```

???
JSON: list of key-value pairs
you cultivate one lock-file per project. as soon as you feel you need to require another lockfile (e.g., due to differing package versions) you should rather start a new R project.

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv/library`

`renv` links your project to a **global package cache** where all your downloaded packages are stored by `renv`. You can think of it as a repository that contains all package versions ever installed on your machine.

.pull-left[
```{r, echo=F, out.width='100%', out.height='100%'}
knitr::include_graphics("https://kevinushey-2020-rstudio-conf.netlify.app/img/package-cache.svg")
```
]
.pull-right[
This approach avoids redundancies, if the same package version is used by multiple projects.

In this case, the installation of the respective package is stored in the global cache with `renv` simply creating a link to this cache within your project. This is what your local **project library** under the path `renv/library` is all about.
]
.pull-right[
.footnote[
Picture by [Kevin Ushey](https://kevinushey-2020-rstudio-conf.netlify.app/slides.html#1).
]]

???
- this approach saves space on disk and makes installing and restoring packages much faster
- renv only copies packages if the linking approach fails

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv/library`

Finally, if you now check your *active library path*, you can observe a new project-specific path (as well as an additional system library path which you need not consider further at this point).
```{r, eval=F}
.libPaths()
```
```
> [1] "C:/Users/.../01_structuring_data_projects/renv/library/R-4.0/x86_64-w64-mingw32"
> [2] "C:/Users/.../AppData/Local/Temp/Rtmp0aVyVr/renv-system-library"
```

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv::snapshot()`

After having successfully set up our project-specific library, let's consider the two other main features of the `renv` package: `renv::snapshot()` and `renv::restore()`.

**`renv::snapshot()`** updates the lockfile by writing the current state of the project library to the lockfile.
```{r, eval=F}
renv::snapshot()
```

--

```
> The following package(s) will be updated in the lockfile:
> 
> # CRAN ===============================
> - glue        [1.4.0 -> 1.4.1]
> - jsonlite    [1.6.0 -> 1.7.0]
> - rmarkdown   [* -> 2.3]
> 
> Do you want to proceed? [y/N]: y
>  * Lockfile written to 'C:/Users/.../01_structuring_data_projects/renv.lock'.
```

---

## 5 `renv`: Project Environments
### 5.3 The `renv` Workflow: `renv::restore()`

After having successfully set up our project-specific library, let's consider the two other main features of the `renv` package: `renv::snapshot()` and `renv::restore()`.

**renv::restore()** updates the project library by restoring the state of the project library as specified in the lockfile.
```{r, eval=F}
renv::restore()
```
--
```
> The following package(s) will be updated:
> 
> # CRAN ===============================
> - rmarkdown   [* -> 2.3]
> - stringr     [* -> 1.4.0]
> 
> Do you want to proceed? [y/N]: y
> Installing stringr [1.4.0] ... OK (copied cache)
> Installing rmarkdown [2.3] ... OK (copied cache)
```

---

## 5 `renv`: Project Environments
### 5.4 Additional Remarks

A project is always bound to a particular version of `renv`. Updating the current version of the `renv` package should be done using the `renv::upgrade()` command.

When setting up `renv` for an existing project, it crawls all project scripts and searches for packages that are mentioned in your code via `renv::dependencies()`. For this presentation, I relied, among others, on the following packages (note that packages are only identified if mentioned in the code by namespace):
```{r, eval=F}
head(renv::dependencies(), 5)
```
```
> Finding R package dependencies ... Done!

>                                          Source          Package
> 1 C:/Users/s_scho53/.../01_rstudio_projects.Rmd        rmarkdown
> 2 C:/Users/s_scho53/.../01_rstudio_projects.Rmd            knitr
> 3 C:/Users/s_scho53/.../01_rstudio_projects.Rmd            knitr
> 4 C:/Users/s_scho53/.../01_rstudio_projects.Rmd            knitr
> 5      C:/Users/s_scho53/.../02_r_notebooks.Rmd        rmarkdown
```

???
Here, only the top 5 entries. `knitr` is three times, according to how often I use the package in my code (here three times for including pictures and gifs in the slides).
